Stochastic Blockmodels for Relational Event Data

### 2012-05-30 docs/submission.tex
Paper edits.
Edited the intro a bit (3rd para mainly)

\title{Stochastic Blockmodels for Relational Event Data}

@@

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.  
For continuous-time network data, recent methods model the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.  
For static networks, methods such as stochastic blockmodels often account for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.  
We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.  
The method is illustrated using dyadic interaction data such as email and Twitter direct messages.  
Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.  
The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.  
Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

@@

The above continuous-time network models assume all nodes in the network behave according to identical dynamics.  
In many cases this is unrealistic.  
For example, in a small organization comprised of several teams, each team may uncover different patterns for collaborating via email --- individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that assumes latent clusters of nodes share similar patterns of interaction.  
The behavior of the model is illustrated with simulated data.  
We describe parameter estimation and the learning of the latent cluster assignments via MCMC.  
Using data sets of dyadic communication among people, we compare the predictive performance of the fitted models to standard baselines.  
Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, sometimes identifying particular \emph{roles} of communication.

@@

\subsection{Exploratory analysis}

@@

\subsubsection*{Acknowledgements}

>>>>

\title{Stochastic Blockmodels for Relational Event Data}
% Latent Variable Modeling of Relational Event Dynamics
% Latent Variable Modeling of Relational Event Sequences
% Latent Variable Modeling for Relational Event Sequences

@@

\begin{abstract}
Observing network data as a stream of relational events provides an opportunity to understand differences in how nodes interact given the past.
For continuous-time network data, a number of approaches have recently been proposed for modeling the rate of dyadic events conditioned on the observed history of events and covariates about the nodes.
For static networks, methods such as stochastic blockmodels have traditionally been used for node-level heterogeneity by assuming latent groups of individuals have similar tendencies in their group-wise interactions.
We propose to combine these two approaches by modeling the event dynamics within and between clusters of nodes.
The method is illustrated using dyadic interaction data such as email communication and Twitter direct messages.
Parameter estimates from the model clearly reveal heterogeneity in the dynamics among groups of individuals.
The fitted models have better predictive accuracy than baselines and relational event models without the latent structure.
Our approach illustrates the efficacy of combining a detailed model for local dependencies and a latent variable model for meso-scale dependencies.
\end{abstract}

@@

A limitation of all of these approaches is the assumption of a single common behavior for all individuals.
In contrast, in real-world networks it is reasonable to expect {\it heterogeneity} in dynamic behavior.  
In an organization, the manner in which an individual communicates with others may be a function of an individual's {\it role} in the organization.
For example, in a university,  email communication patterns over time between professors, students, and staff will likely be quite different within and between the three groups.
%PS: not totally happy with this example since the "roles" are completely determined by one's "job title" and I think we would prefer something
% less explicit, e.g., that there are "leaders", "followers", etc, that go beyond just job titles. So feel free to edit this further!

%
%comprised of several teams, each team may uncover different patterns for collaborating via email---individuals in one group may respond more quickly, while another group may preferentially send to highly active individuals.

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data to capture node-level heterogeneity, where we assume that latent clusters of nodes share similar patterns of interaction.
We describe parameter estimation and learning of  latent cluster assignments via MCMC, and illustrate the behavior of the model  with simulated data.
Using several real-world social network data sets involving dyadic communication, we compare the predictive performance of the fitted models to standard baselines.
Finally we show that the parameter estimates exhibit an interpretable structure to the event dynamics, allowing identification of particular \emph{roles} in communication.

@@

\subsection{Model-based exploratory analysis}

@@

%\subsubsection*{Acknowledgements}
